Empirical Evidence that Modern Byte Pair Tokenizers are Zipfian

### 2025-05-31 writeup/paper.tex
ok thats done yay

\begin{document}
\maketitle
\begin{abstract}
A majority of large language models ingest word fragments produced by an algorithm known as byte pair encoding. This algorithm groups high-frequency element pairings in natural language into individual units called \textit{tokens}. A natural question is whether the frequency distribution of the tokens produced by this pairing process deviate significantly from the source language's. Zipf showed that many natural language's frequency distribution follows a power law, commonly known as Zipf's Law. We examine two modern tokenizer's adherence to Zipf's law at the token level. We provide empirical evidence that these tokenizers are Zipfian on two corpuses, and speculate as to why this is. Additionally, we give evidence that Zipfness is preserved over many individual steps of byte pair encoding.
\end{abstract}

@@

George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurences of a word, the word's \textit{frequency}, in many human languages exhibit the same power law distribution. Specifically, Zipf said the work frequency in a corpus is proportional to the word's frequency rank, that being

@@

The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}. Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted. This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots. This transition is found around the $10^4$\textit{-th} ranked word \cite{cancho2000} \cite{yu2018zipfslaw50languages}. The authors further explore the linguistic relevance and universality of these multiple trends of words.

Nearly a century after Zipf's discovery, large language models (LLMs) generate text comparable to human communication \cite{llmturingtest}. LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}. The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} introduced by Gage \cite{bpegage} for file compression. An adjustment of BPE was popluarized for natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}, wherein instead of seeking to optimally compress a corpus of text, it instead seeks to map a corpus to a small fixed vocabulary. In this paper, we consider the natural language processing variant of the algorithm. For BPE, the \textit{elements} of a corpus are what portion of the text is paired in the first iteration of the algorithm. Elements can be words, phonemes, or letters. Modern tokenizers use the bytes of the Unicode representation as the elements, and so are called \textit{byte-level} tokenizers \cite{radford2019language}.

@@

We compute the byte, word, and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020). We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare the fit to fitted exponential and log-normal distributions. We do not remove any words or tokens from consideration unless otherwise stated.

Additionally, we create a synthetic dataset of $10^5$ samples from a Zipf distribution alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.

@@

We find that token vocabularies are strongly Zipfian, excluding extremely rare tokens. This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.

For bookscorpus, we find 1.3M distinct words, 31,729 distinct RoBERTa tokens, and 12,859 distinct GPT-4o tokens. For MiniPile, we find 32M distinct words, 50,165 distinct RoBERTa tokens, and 178,416 distinct GPT-4o tokens.

We plot the log-log rank frequencies for each corpus in Figure \ref{fig:experiments}. We find a clear linear trend over a large number of words and tokens. We only see deviations from the Zipfian trend for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4o tokenizer for the MiniPile corpus. However, the GPT-4o tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section:intro}.

@@

We also report the lowest rank words and tokens for each tokenizer and corpus to identify trends across models and corpuses. We report these words and tokens both including and excluding control tokens and punctuation, which are summarized in Table \ref{tbl:token ranks} and Table \ref{tbl:token ranks no control} respectively.

\begin{table*}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Word}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4o}}  \\
        \hline
        1 & . & the & <s> & ␣ & `` & , \\
        \hline
        2 & , & of & \textbackslash n & \textbackslash n & i & ␣the \\
        \hline
        3 & the & and & <\textbackslash s> & . & he & . \\
        \hline
        -1 & restain & RootDir, & seq & ␣Archdemon & wares & aryny \\
        \hline
        -1 & liarliar & homocystinemia & okemon & ␣petertodd & slan & verlening \\
        \hline
        -1 & shop-that & halfday & ython & ␣councill & wier & ␣myx \\
        \hline
    \end{tabular}
    \caption{Summary of most and least common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, \textit{including} control tokens and punctuation. A rank of -1 indicates a word or token that appears only once in the given corpus. The space character (Unicode U+0020) is denoted by the `␣' character. }
    \label{tbl:token ranks}
\end{table*}

@@

We report the Kolmogorov-Smirnov test statistics for fitted Zipf-Mandelbrot, Exponential, and log-normal distributions on $10^5$ samples from the rank-frequency data in Table \ref{tbl:kstest}.

\begin{table*}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4o}}  \\
        \hline
        Zipf & 0.3324 & 03136 & 0.3307 & 0.3157 & 0.3785 & 0.3136 \\
        \hline
        Exponential & 0.5457 & 0.4405 & 0.5180 & 0.4728 & 0.5042 & 0.4287 \\
        \hline
        log-normal & 0.3708 & 0.3129 & 0.3675 & 0.3581 & 0.4230 & 0.2544 \\
        \hline
    \end{tabular}
    \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
    \label{tbl:kstest}
\end{table*}

Figure \ref{fig:theoryexperiments} shows the mean squared error (MSE) between the Zipfian prediction and successive applications of a single step of BPE over the Zipfian synthetic dataset.

@@

We demonstrate a strong Zipfian trend in all tokenizers and corpuses studied, regardless of whether or not the tokenizer was trained on a given corpus. Deviations from the Zipf prediction only appear for the final few tokens. Over the token vocabulary, the most common tokens tend to be punctuation, newline characters, and control tokens. Excluding punctuation and special tokens, we recover the known most common English words.

Additionally, synthetic data that is generated from a Zipfian distribution remain Zipfian after many successive steps of BPE.

@@

The BPE algorithm, despite building vocabularies using the most frequent pairings of elements, generates vocabularies that are still Zipfian, up to the rarest few tokens.

Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}. Most significant was Belevitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian. This indicates that Zipf's law may be due to the rank ordering of words more than the underlying formation of language, which is what Zipf was originally interested in. Can Belevitch's proof be shown to be invariant for some number of byte pair encoding transformations?

>>>>

\begin{document}
\maketitle
\begin{abstract}
A majority of large language models ingest word fragments produced by an algorithm known as byte pair encoding. This algorithm groups high-frequency element pairings in natural language into individual units called \textit{tokens}. A natural question is whether the frequency distribution of the tokens produced by this pairing deviate significantly from the source language's. Zipf showed that many natural language's frequency distribution follow a power law, commonly known as Zipf's Law. We examine two modern tokenizer's adherence to Zipf's law at the token level. We provide empirical evidence that these tokenizers are Zipfian on two corpuses, and speculate as to why this is. Additionally, we give evidence that Zipfness is preserved over many individual steps of byte pair encoding.
\end{abstract}

@@

George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurences of a word, the word's \textit{frequency}, exhibits a power law. Specifically, Zipf said the word frequency in a corpus is proportional to the word's frequency rank, that being

@@

The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}. Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted. This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots. In English text, this transition is found around the $10^4$\textit{-th} ranked word \cite{cancho2000} \cite{yu2018zipfslaw50languages}. The authors further explore the linguistic relevance and universality of these multiple trends.

Nearly a century after Zipf's discovery, large language models (LLMs) generate text comparable to human communication \cite{llmturingtest}. LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}. The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding algorithm (BPE)}. BPE was introduced by Gage \cite{bpegage} for file compression, and later adapted for use in natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}. The NLP version of BPE, instead of seeking to optimally compress a corpus of text into a vocabulary of any size, instead maps a corpus to a small fixed vocabulary. In this paper, we consider the natural language processing variant of the algorithm.

BPE also requires specifying the \textit{elements} of a corpus, which are the portion of the text that is considered a indivisible chunk of the text. Elements can be words, phonemes, or letters. Modern tokenizers use the bytes of the Unicode representation as the elements, and so are called \textit{byte-level} BPE tokenizers \cite{radford2019language}.

@@

We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020). We then compute the data's $\chi^2$ goodness-of-fit statistic for the fitted Zipf distribution, and compare the fit to exponential and log-normal distributions. We do not remove any words or tokens from consideration unless otherwise stated.

Additionally, we create a synthetic dataset of $10^5$ samples from a zipf-distibuted alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.

@@

For bookscorpus, we find 1.3M distinct words, 31,729 distinct RoBERTa tokens, and 12,859 distinct GPT-4o tokens. For MiniPile, we find 32M distinct words, 50,165 distinct RoBERTa tokens, and 178,416 distinct GPT-4o tokens. We plot the log-log rank frequency for both corpuses in Figure \ref{fig:experiments}.

@@

We also report the least and most common words and tokens for each tokenizer and corpus. We report these words and tokens both including and excluding control tokens and punctuation, which are summarized in Table \ref{tbl:token ranks} and Table \ref{tbl:token ranks no control}.

\begin{table*}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Word}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4o}}  \\
        \hline
        1 & . & the & <s> & ␣ & `` & , \\
        \hline
        2 & , & of & \textbackslash n & \textbackslash n & i & ␣the \\
        \hline
        3 & the & and & <\textbackslash s> & . & he & . \\
        \hline
        -1 & restain & RootDir, & seq & ␣Archdemon & wares & aryny \\
        \hline
        -1 & liarliar & homocystinemia & okemon & ␣petertodd & slan & verlening \\
        \hline
        -1 & shop-that & halfday & ython & ␣councill & wier & ␣myx \\
        \hline
    \end{tabular}
    \caption{Summary of most and least common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, \textit{including} control tokens and punctuation. A rank of -1 indicates a word or token that appears only once in the given corpus. The space character (Unicode U+0020) is denoted by the `␣' character.}
    \label{tbl:token ranks}
\end{table*}

@@

We report the $\chi^2$ test statistics for fitted Zipf-Mandelbrot in Table \ref{tbl:kstest}, where we use Equation \ref{eq:zipfmandelbrot} with $a=1.1$ and $b=0$ for all Zipf tests.

\begin{table*}
    \centering
    \begin{tabular}{|c|c|c|c|}
        \hline
        \textbf{Distribution} & \textbf{Words} & \textbf{RoBERTa} & \textbf{GPT-4o}  \\
        \hline
        bookscorpus & $2.5 \cdot 10^{9}$ & $1.6 \cdot 10^{9}$ & $1.4 \cdot 10^{8}$  \\
        \hline
        MiniPile &  $1.4 \cdot 10^{9}$ & $4.1 \cdot 10^{9}$ & $1.8 \cdot 10^{9}$ \\
        \hline
    \end{tabular}
    \caption{$\chi^2$ test statistics for distribution fit for each tokenizer and corpus, each of which have a p value of near $0$}
    \label{tbl:kstest}
\end{table*}

Finally, we compute the mean squared error (MSE) between the Zipfian prediction and successive applications of a single step of BPE over the Zipfian synthetic dataset in Figure \ref{fig:theoryexperiments}.

@@

In general, we find strong evidence that BPE tokenizers are Zipfian. Deviations from the Zipfian trend only appear for the final few RoBERTa tokens for both corpuses and the GPT-4o tokens for the MiniPile corpus. This includes in and out-of-distribution text. However, the GPT-4o tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section:intro}. We speculate that this may be due to Minipile being a multilingual dataset, which may affect the the distribution for a multilingual tokenizer such as GPT-4o. The $\chi^2$ goodness-of-fit tests confirm a Zipf describes the data well, all having p-values $<<0.0001$.

Over the token vocabulary, the most common tokens tend to be punctuation, newline characters, and control tokens. Excluding punctuation and special tokens, we recover the known most common English words.

Additionally, applying the BPE pairing process to the synthetic data shows almost no deviations from the original Zipf distribution for hundreds of iterations. However, after about $700$ iterations the MSE from Zipf does begin to increase, though the error is still relatively small.

@@

The byte pair encoding algorithm, despite building vocabularies using the most frequent pairings of elements, generates vocabularies that are Zipfian, up to the rarest few tokens. We demonstate this over multiple tokenizers and corpuses, including both in and out-of-distribution text. We also provide empirical evidence that Zipfness is preserved over many iterations of BPE.

Noticeable theoretical work has been done to explain Zipf's law for language \cite{li1992}. Most significant was Belevitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian. This indicates that Zipf's law may be due to the rank ordering of words more than the underlying formation of language. Can Belevitch's proof be shown to be invariant for some number of byte pair encoding transformations?
